---
title: "Workshop on 'Applications for artificial intelligence (AI) in risk assessment' (AiRA)"
author: ""
site: bookdown::bookdown_site
output: 
    bookdown::gitbook:
        css: style.css
        number_sections: false
        anchor_sections: false
        split_by: chapter
        config:
            sharing:
                 github: yes
                 facebook: no
                 twitter: no
                 all: no
            toc:
                collapse: section
                scroll_highlight: yes
                before: <li class="toc-logo"><a href="./"></a> <h4 class=".paddingtitel ">AI in Risk Assessment (AiRA) </h2></li>
header-includes:
  - \usepackage{fontawesome5}
---

Download the source code [`r icons::fontawesome("github")`](https://github.com/ontox-hu/aspis-workshop-ai-tox)


```{r include = FALSE}
# Define the web addresses for the course
# run as needed, not included automatically because of stuff (aka git complains)
githuburl <- "https://github.com/ontox-hu/aspis-workshop-ai-tox"
readerurl <- "https://rstudio-connect.hu.nl/aira/"
email <- "marc.teunis@hu.nl"

save(
    list = c("githuburl", "readerurl", "email"), 
    file = "course_urls.RData")
```


```{r include = FALSE, eval = TRUE}
# set CSS for objects
knitr::opts_chunk$set(
  class.source="Rchunk", 
  class.output="Rout", 
  warning = FALSE,
  error = FALSE,
  message = FALSE)
```

# Introduction {-}

[These workshop materials was produced for the ASPIS/ESTIV Conference, 2022, to be held on 25 November in Sitges (Barcelona) in Spain](https://estiv.org/congress2022/)

```{r, echo=FALSE}
knitr::include_graphics(here::here(
  "images",
  "estiv-banner.png"
))
```

## Disclaimer on this work
We would like to stress that the code and work included in this repo and worshop is experimental. It was not reviewed by a peer assessment process. The code, examples and work should be considered accordingly. The work is meant for illustrative and educational purposes. The authors do not take any responsiblility for the use, re-use, application or derivates from the work included in this repository. We appreciate attribution of the original work and adapted work by the authors of this repo.

## AiRA
Welcome to the workshop on "Applications for artificial intelligence (AI) in risk assessment" (AiRA), Friday 25 November, from 1 to 4 pm, during the ESTIV/ASPIS Conference in Sitges, Barcelona in Spain. This Workshop consists of 4 learning activities:

 1. An introduction to R and the tidyverse
 1. An introduction to Tidymodels in R (adapted from the Tidymodels documentation)
 1. An introduction to Read Accross, based on Compound similarity
 1. An introduction to Machine Learning approaches for classification of chemicals

If you are already acquainted with R and a practiced user of the `{tidyverse}` suit of R packages, you can skip activity 1. 
If you already are familiar with the `{tidymodels}` workflow, you might want to skip or only glimpse over activity 2.


You can find these lessons in the menu on the left. Each lesson starts with a short introduction by the teachers. After that, you are supposed to study the lessons yourself and make the exercises. During the lessons, you can ask questions to the teachers and the teachers will provide feedback. 

Please note: this is a crash course. We scheduled this course to allow you to throw yourself into R and be able to see yourself making progress very quickly. 

We wish you good luck and we hope you will have a lot of fun with R!

## Course material

The course material consists of the following:

- Instruction pages (menu to the left). 
- Exercises.

To navigate through the different materials, you can use the menu on the left. The course is based on the online book [R for Data Science](https://r4ds.had.co.nz/index.html).


## Exercises

The exercises consist of writing R code to solve data analysis problems. Answers are available for each exercise. Most likely you will not finish all exercises during the workshop. The best way to learn programming is to practice, practice and practice by making the exercises. 

## R and RStudio

During the course, we will use R in the Integrated Development Environment  **RStudio**, which you can download and install for your system using the links below:

Download R [here](https://cran.r-project.org/)
Download RStudio [here](https://posit.co/download/rstudio-desktop/)

## Resources and Bookdown
This website has been created using the `{bookdown}` R package
```{r}
citation(package = "bookdown")
```
I you want to learn more on bookdown, see [this link](https://bookdown.org/)

## Learning objectives

After this course, you will be able to 

- use R to perform a structure-based Generalised Read Across
- explain the basics of machine learning and Deep learning in toxicology.
- Build a Neural Network and monitor its performance in R
- Build a classical Machine Learning model and evaluate its performance
- Run model-tuning to find optimal hyperparameters 

## Reference and original source
Some of the materials presented here were originally developed by Dr. Grace Patlewicz and Dr. Julia E. Rager as part of [TAME](https://uncsrp.github.io/Data-Analysis-Training-Modules/machine-learning-and-predictive-modeling.html#machine-learning-and-predictive-modeling), and was reproduced and adapted with permission of the authors. See for the publications and [the complete Toolbox:](https://github.com/UNCSRP/Data-Analysis-Training-Modules). This toolbox is a good place to start for Introductory Data Science, Chemical-Biological Analyses, Predictive Modeling, and Database Mining for Environmental Health Research. 

## Attribution

This work is distributed under a CC BY-NC 4.0 licence. Please cite this work as:

Corradi, M.; De Haan, A. & Teunis, M.A.T., 2022, "Workshop on 'Applications for artificial intelligence (AI) in risk assessment' (AiRA)", ASPIS Cluster Meeting, Sitges, Barcelona, https://github.com/ontox-hu/aira 

and include the original source as:

Roell K, Koval LE, Boyles R, Patlewicz G, Ring C, Rider CV, Ward-Caviness C, Reif DM, Jaspers I, Fry RC, Rager JE. Development of the InTelligence And Machine LEarning (TAME) Toolkit for Introductory Data Science, Chemical-Biological Analyses, Predictive Modeling, and Database Mining for Environmental Health Research. Front Toxicol. 2022 Jun 22;4:893924. doi: 10.3389/ftox.2022.893924. PMID: 35812168; PMCID: PMC9257219.
https://doi.org/10.3389/ftox.2022.893924

Some materials were adapted from: [TAME](https://uncsrp.github.io/Data-Analysis-Training-Modules/machine-learning-and-predictive-modeling.html#machine-learning-and-predictive-modeling), and was reproduced and adapted with permission of the authors. See for the publications and [the complete Toolbox:](https://github.com/UNCSRP/Data-Analysis-Training-Modules). This toolbox is a good place to start for Introductory Data Science, Chemical-Biological Analyses, Predictive Modeling, and Database Mining for Environmental Health Research. 



